One page description

### 2016-06-30 publications/Journal/1-introduction.tex
Consistent italics and sans-serif fonts

This paper introduces DEVS-Ex-Machina\footnote{\url{https://bitbucket.org/bcardoen/devs-ex-machina}} (``dxex''): our simulation tool~\cite{dxex} which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
The selected synchronization protocol is transparent to the simulated model: users only determine which protocol to use.
Users who simulate a wide variety of models, with different ideal synchronization protocols, can run the same model with different synchronization protocols.
Since the ideal synchronization protocol might change throughout the simulation, we also support runtime switching between these protocols.
This runtime switching can be based on performance metrics, which are logged during simulation.
Information is made available to a separate component, where a choice can be made about which synchronization protocol to use next. % It's manual now, can be extended to full automatic.
Additionally, we investigate in this paper how model allocation influences the performance of our synchronization protocols.
To this end, we have included an allocation component to our simulation kernel.

Our tool is based on PythonPDEVS, but implemented in C++11 for increased performance, using features from the new C++14 standard when supported by the compiler.
Unlike PythonPDEVS, dxex only supports multicore parallelism, thus no distributed simulation.

We implemented a model that, depending on a single parameter, changes its ideal synchronization protocol.
Using several models, we demonstrate the factors influencing the performance under a given synchronization protocol.
Dxex, then, is used to compare simulation using exactly the same tool, but with a varying synchronization protocol.
With dxex users can always opt to use the fastest protocol available, and through its modularity, users could even implement their own.
To verify that this modularity does not hinder performance, we compare to another Parallel DEVS simulation kernel, called adevs~\cite{adevs}.

%Usage of Hyperref with hardcoded label is hardly ideal, but ~\ref doesn't work if the layout disables numbered sections, so a hack is needed. The alternative is \nameref but that isn't all that readable.
The remainder of this paper is organized as follows:
Section~\textsc{\nameref{sec:2-background}} introduces the necessary background on synchronization protocols.
Section~\textsc{\nameref{sec:3-features}} elaborates on our design that enables this flexibility.
In Section~\textsc{\nameref{sec:4-performance}}, we evaluate performance of our tool by comparing its different synchronization protocols, and compare to adevs.
We continue by introducing runtime switching of synchronization protocols and different options for model allocation in Section~\textsc{\nameref{sec:4b-hotswap}} and Section~\textsc{\nameref{sec:4a-allocation}}, respectively.
Related work is discussed in Section~\textsc{\nameref{sec:5-related-work}}.
Section~\textsc{\nameref{sec:6-conclusion}} concludes the paper and gives future work.

>>>>

This paper introduces \textit{DEVS-Ex-Machina}\footnote{\url{https://bitbucket.org/bcardoen/devs-ex-machina}} (``\textit{dxex}''): our simulation tool~\cite{dxex} which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
The selected synchronization protocol is transparent to the simulated model: users only determine which protocol to use.
Users who simulate a wide variety of models, with different ideal synchronization protocols, can run the same model with different synchronization protocols.
Since the ideal synchronization protocol might change throughout the simulation, we also support runtime switching between these protocols.
This runtime switching can be based on performance metrics, which are logged during simulation.
Information is made available to a separate component, where a choice can be made about which synchronization protocol to use next. % It's manual now, can be extended to full automatic.
Additionally, we investigate in this paper how model allocation influences the performance of our synchronization protocols.
To this end, we have included an allocation component to our simulation kernel.

Our tool is based on \textit{PythonPDEVS}, but implemented in C++11 for increased performance, using features from the new C++14 standard when supported by the compiler.
Unlike \textit{PythonPDEVS}, \textit{dxex} only supports multicore parallelism, thus no distributed simulation.

We implemented a model that, depending on a single parameter, changes its ideal synchronization protocol.
Using several models, we demonstrate the factors influencing the performance under a given synchronization protocol.
\textit{Dxex}, then, is used to compare simulation using exactly the same tool, but with a varying synchronization protocol.
With \textit{dxex} users can always opt to use the fastest protocol available, and through its modularity, users could even implement their own.
To verify that this modularity does not hinder performance, we compare to another \textsf{Parallel DEVS} simulation kernel, called \textit{adevs}~\cite{adevs}.

%Usage of Hyperref with hardcoded label is hardly ideal, but ~\ref doesn't work if the layout disables numbered sections, so a hack is needed. The alternative is \nameref but that isn't all that readable.
The remainder of this paper is organized as follows:
Section~\textsc{\nameref{sec:2-background}} introduces the necessary background on synchronization protocols.
Section~\textsc{\nameref{sec:3-features}} elaborates on our design that enables this flexibility.
In Section~\textsc{\nameref{sec:4-performance}}, we evaluate performance of our tool by comparing its different synchronization protocols, and compare to \textit{adevs}.
We continue by introducing runtime switching of synchronization protocols and different options for model allocation in Section~\textsc{\nameref{sec:4b-hotswap}} and Section~\textsc{\nameref{sec:4a-allocation}}, respectively.
Related work is discussed in Section~\textsc{\nameref{sec:5-related-work}}.
Section~\textsc{\nameref{sec:6-conclusion}} concludes the paper and gives future work.

### 2016-06-30 publications/Journal/3-features.tex
Consistent italics and sans-serif fonts

Historically, dxex is based on PythonPDEVS~\cite{PythonPDEVS}.
Python is a good language for prototypes, but performance has proven insufficient to compete with other simulation kernels~\cite{MasterThesis}.
Dxex is a C++11-based \textsf{Parallel DEVS} simulator, based on the design of PythonPDEVS.
Whereas the feature set is not too comparable, the architectural design, simulation algorithms, and optimizations, are highly similar.

We will not make a detailed comparison with PythonPDEVS here, but only mention some supported features.
Dxex supports, similarly to PythonPDEVS, the following features: direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions~\cite{JDF}, and a modular tracing and scheduling framework~\cite{PythonPDEVS}.
We do not elaborate on these features in this paper.
But whereas PythonPDEVS only supports optimistic synchronization, dxex support multiple synchronization protocols (though only in parallel).
This is in line with the design principle of PythonPDEVS: allow users to pass performance hints to the simulation kernel.
In our case, a user can pass the simulation kernel the synchronization protocol to use for this model, or even switch the synchronization protocol during runtime.
Our implementation in C++11 also allows for optimizations which were plainly impossible in an interpreted language. Dxex will use new optimizations from the C++14 standard when possible.

Since there is no universal \textsf{DEVS} model standard, dxex models are incompatible with PythonPDEVS and vice versa.
This is due to dxex models being grafted on C++11, whereas PythonPDEVS models are grafted on Python.

@@

\subsection{Synchronization protocols}
We previously explained the existence of different synchronization protocols, each optimized for a specific kind of model.
As no single synchronization protocol is ideal for all models, a general purpose simulation tool should support multiple protocols.
Currently, most parallel simulation tools choose only a single synchronization protocol due to the inherent differences between protocols.
We argue that a real general purpose simulation tool should support sequential, conservative, and optimistic synchronization, as is the case for dxex.

@@

\subsubsection{Sequential}
Our sequential simulation algorithm is very similar to the one found in PythonPDEVS, including many optimizations.
Minor modifications were made, though, such that it can be overloaded by different synchronization protocol implementations.
This way, the \textsf{DEVS} simulation algorithm is implemented once, but parts can be overridden as needed.
In theory, more synchronization protocols (\textit{e.g.}, other algorithms for conservative synchronization) can be added without altering our design.

An overview of dxex's design is given in Figure~\ref{fig:class_diagram}.
It shows that there is a simulation \texttt{Core}, which simulates the \texttt{AtomicModel}s connected to it.
The superclass \texttt{Core} represents the sequential simulation core.
Subclasses define specific variants, such as \texttt{ConservativeCore} (conservative synchronization), \texttt{OptimisticCore} (optimistic synchronization), and \texttt{DynamicCore} (\textsf{Dynamic Structure DEVS}).

\begin{figure*}
    \includegraphics[width=\textwidth]{fig/cores_class_diagram.eps}
	\caption{Dxex design.}
	\label{fig:class_diagram}
\end{figure*}

@@

\subsubsection{Memory Management}
\label{sec:4-subsec:overhead-pgraph:memory}
\begin{figure}
    \center
    \includegraphics[width=\columnwidth]{fig/memory_allocators_parallel.eps}
    \caption{Effect of memory allocators on parallel execution time.}
    \label{fig:memallocators_parallel}
\end{figure}
Memory management is traditionally seen as one of the major bottlenecks in parallel computation~\cite{Memory}: memory bandwidth doesn't increase as fast as the number of cores using it.
While this is always a problem, it is aggravated in dxex by providing automatic memory management for events and states.
A model written for sequential simulation will run correctly in a conservative or optimistic simulation without altering (from the point of view of the modeller) the (de)allocation semantics of events or states.

@@

Dxex uses Boost Pool~\cite{boostpool} allocators in parallel simulation kernels and arena-style allocators for sequential simulation.
The latter can be faster, but at the cost of extra configuration.
The allocators are supplemented by the library \textit{tcmalloc}~\cite{tcmalloc}, which reduces lock contention in \texttt{malloc} calls.

We primarily investigate this for optimistic simulation, as this is the most memory consuming mode of simulation~\cite{FujimotoBook}.
Simulation execution times for all four combinations are shown in Figure~\ref{fig:memallocators_parallel}.
Optimistic simulation greatly benefits from the use of \textit{tcmalloc}, regardless of the allocator.
Nonetheless the pool allocator also reduces the allocation overhead, though only by a relatively small fraction.
Both techniques are required to reduce the overhead of memory allocations in dxex, and on by default.

@@

We of course still need to guarantee determinism and isolation between the calls to the RNG, as well as avoiding excessive synchronization.
Dxex uses the Tina RNG collection (TRNG)~\cite{PhysRevE.75.066701} as an alternative random number generator with performance and multithreading in mind.
Since the RNG is an implicit part of the state in the DEVS formalism, though often not implemented as such, we evaluated performance for both approaches: one global RNG per thread, and one RNG per atomic DEVS model.

We see in Figure~\ref{fig:Queuerngspeedup} that storing the RNG in the state is very expensive for the default STL random number generator.
This is primarily caused by the significant difference in size: 2504 bytes for the STL random number generator, and 24 bytes for the Tina random number generator.
Dxex's sequential and conservative kernels are insensitive to storing the RNG object in the atomic model state, since no copying/state saving occurs in dxex conservative simulation.
The optimistic kernel is clearly affected, as it needs to copy more bytes in every transition due to state saving.

Figure~\ref{fig:Queuerngspeedup} shows that dxex in sequential simulation gets three times faster by using TRNG, than when using the STL RNG.
For parallel simulation, the synchronization overhead seems to be the main bottleneck, as seen by the big speedup gap between sequential and parallel simulation.
Conservative synchronization is almost insensitive to the changing of the RNG, though a slight increase in performance can be noted.
Optimistic synchronization gets much slower when the RNG becomes part of the model state, since the state needs to be copied as well.
This becomes a significant overhead when using the STL RNG, since performance plummets to a fraction of the original.
Using TRNG avoids this problem completely, as the size of the RNG state is negligible.

>>>>

Historically, \textit{dxex} is based on \textit{PythonPDEVS}~\cite{PythonPDEVS}.
Python is a good language for prototypes, but performance has proven insufficient to compete with other simulation kernels~\cite{MasterThesis}.
\textit{Dxex} is a C++11-based \textsf{Parallel DEVS} simulator, based on the design of \textit{PythonPDEVS}.
Whereas the feature set is not too comparable, the architectural design, simulation algorithms, and optimizations, are highly similar.

We will not make a detailed comparison with \textit{PythonPDEVS} here, but only mention some supported features.
\textit{Dxex} supports, similarly to \textit{PythonPDEVS}, the following features: direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions~\cite{JDF}, and a modular tracing and scheduling framework~\cite{PythonPDEVS}.
We do not elaborate on these features in this paper.
But whereas \textit{PythonPDEVS} only supports optimistic synchronization, \textit{dxex} support multiple synchronization protocols (though only in parallel).
This is in line with the design principle of \textit{PythonPDEVS}: allow users to pass performance hints to the simulation kernel.
In our case, a user can pass the simulation kernel the synchronization protocol to use for this model, or even switch the synchronization protocol during runtime.
Our implementation in C++11 also allows for optimizations which were plainly impossible in an interpreted language. \textit{Dxex} will use new optimizations from the C++14 standard when possible.

Since there is no universal \textsf{DEVS} model standard, \textit{dxex} models are incompatible with \textit{PythonPDEVS} and vice versa.
This is due to \textit{dxex} models being grafted on C++11, whereas \textit{PythonPDEVS} models are grafted on Python.

@@

\subsection{Synchronization protocols}
We previously explained the existence of different synchronization protocols, each optimized for a specific kind of model.
As no single synchronization protocol is ideal for all models, a general purpose simulation tool should support multiple protocols.
Currently, most parallel simulation tools choose only a single synchronization protocol due to the inherent differences between protocols.
We argue that a real general purpose simulation tool should support sequential, conservative, and optimistic synchronization, as is the case for \textit{dxex}.

@@

\subsubsection{Sequential}
Our sequential simulation algorithm is very similar to the one found in \textit{PythonPDEVS}, including many optimizations.
Minor modifications were made, though, such that it can be overloaded by different synchronization protocol implementations.
This way, the \textsf{DEVS} simulation algorithm is implemented once, but parts can be overridden as needed.
In theory, more synchronization protocols (\textit{e.g.}, other algorithms for conservative synchronization) can be added without altering our design.

An overview of \textit{dxex}'s design is given in Figure~\ref{fig:class_diagram}.
It shows that there is a simulation \texttt{Core}, which simulates the \texttt{AtomicModel}s connected to it.
The superclass \texttt{Core} represents the sequential simulation core.
Subclasses define specific variants, such as \texttt{ConservativeCore} (conservative synchronization), \texttt{OptimisticCore} (optimistic synchronization), and \texttt{DynamicCore} (\textsf{Dynamic Structure DEVS}).

\begin{figure*}
    \includegraphics[width=\textwidth]{fig/cores_class_diagram.eps}
	\caption{\textit{Dxex} design.}
	\label{fig:class_diagram}
\end{figure*}

@@

\subsubsection{Memory Management}
\label{sec:4-subsec:overhead-pgraph:memory}
\begin{figure}
    \center
    \includegraphics[width=\columnwidth]{fig/memory_allocators_parallel.eps}
    \caption{Effect of memory allocators on parallel execution time.}
    \label{fig:memallocators_parallel}
\end{figure}
Memory management is traditionally seen as one of the major bottlenecks in parallel computation~\cite{Memory}: memory bandwidth doesn't increase as fast as the number of cores using it.
While this is always a problem, it is aggravated in \textit{dxex} by providing automatic memory management for events and states.
A model written for sequential simulation will run correctly in a conservative or optimistic simulation without altering (from the point of view of the modeller) the (de)allocation semantics of events or states.

@@

\textit{Dxex} uses \textit{Boost Pool}~\cite{boostpool} allocators in parallel simulation kernels and arena-style allocators for sequential simulation.
The latter can be faster, but at the cost of extra configuration.
The allocators are supplemented by the library \textit{tcmalloc}~\cite{tcmalloc}, which reduces lock contention in \texttt{malloc} calls.

We primarily investigate this for optimistic simulation, as this is the most memory consuming mode of simulation~\cite{FujimotoBook}.
Simulation execution times for all four combinations are shown in Figure~\ref{fig:memallocators_parallel}.
Optimistic simulation greatly benefits from the use of \textit{tcmalloc}, regardless of the allocator.
Nonetheless the pool allocator also reduces the allocation overhead, though only by a relatively small fraction.
Both techniques are required to reduce the overhead of memory allocations in \textit{dxex}, and are turned on by default.

@@

We of course still need to guarantee determinism and isolation between the calls to the RNG, as well as avoiding excessive synchronization.
\textit{Dxex} uses the Tina RNG collection (TRNG)~\cite{PhysRevE.75.066701} as an alternative random number generator with performance and multithreading in mind.
Since the RNG is an implicit part of the state in the \textsf{Parallel DEVS} formalism, though often not implemented as such, we evaluated performance for both approaches: one global RNG per thread, and one RNG per atomic \textsf{DEVS} model.

We see in Figure~\ref{fig:Queuerngspeedup} that storing the RNG in the state is very expensive for the default STL random number generator.
This is primarily caused by the significant difference in size: 2504 bytes for the STL random number generator, and 24 bytes for the Tina random number generator.
\textit{Dxex}'s sequential and conservative kernels are insensitive to storing the RNG object in the atomic model state, since no copying/state saving occurs in \textit{dxex} conservative simulation.
The optimistic kernel is clearly affected, as it needs to copy more bytes in every transition due to state saving.

Figure~\ref{fig:Queuerngspeedup} shows that \textit{dxex} in sequential simulation gets three times faster by using TRNG, than when using the STL RNG.
For parallel simulation, the synchronization overhead seems to be the main bottleneck, as seen by the big speedup gap between sequential and parallel simulation.
Conservative synchronization is almost insensitive to the changing of the RNG, though a slight increase in performance can be noted.
Optimistic synchronization gets much slower when the RNG becomes part of the model state, since the state needs to be copied as well.
This becomes a significant overhead when using the STL RNG, since performance plummets to a fraction of the original.
Using TRNG avoids this problem completely, as the size of the RNG state is negligible.

### 2016-06-30 publications/Journal/5-hotswap.tex
Consistent italics and sans-serif fonts

Simply because a synchronization protocol is ideal at the start of the simulation, does not mean that it will still be ideal during the simulation.
It is well known, and repeated in the previous section, that model behaviour significantly influences the ideal synchronization protocol.
Contrary to many modelling formalisms, the DEVS formalisms makes it possible to model basically any kind of discrete event model.
As such, it is possible for the model to significantly change its behaviour throughout the simulation.

Defining the ideal synchronization protocol at the start of the simulation, when information about future model behaviour is scarce, might therefore not offer the best possible performance.
In dxex, we not only make it possible to define the synchronization protocol to use, but also to change this decision throughout simulation.
To do this, all kernels are notified of the switch and they are forced to stop simulation.
When stopped, each kernel instantiates a new core, with the new synchronization protocol, that is provided with the simulation state of the previous core.
Simulation is then resumed with the new cores after the previous ones are destroyed.

@@

For example, if the actually seen lookahead is significantly higher than the defined lookahead, it might be interesting to switch to optimistic synchronization.
When the number of rollbacks is excessively high, switching to conservative synchronization might be considered.
And when neither of these two is an option, the simulation might just (temporarily) have to fall back to sequential simulation.

>>>>

Simply because a synchronization protocol is ideal at the start of the simulation, does not mean that it will still be ideal during the simulation.
It is well known, and repeated in the previous section, that model behaviour significantly influences the ideal synchronization protocol.
Contrary to many modelling formalisms, the \textsf{DEVS} formalisms makes it possible to model basically any kind of discrete event model.
As such, it is possible for the model to significantly change its behaviour throughout the simulation.

Defining the ideal synchronization protocol at the start of the simulation, when information about future model behaviour is scarce, might therefore not offer the best possible performance.
In \textit{dxex}, we not only make it possible to define the synchronization protocol to use, but also to change this decision throughout simulation.
To do this, all kernels are notified of the switch and they are forced to stop simulation.
When stopped, each kernel instantiates a new core, with the new synchronization protocol, that is provided with the simulation state of the previous core.
Simulation is then resumed with the new cores after the previous ones are destroyed.

@@

For example, if the actually seen lookahead is significantly higher than the defined lookahead, it might be interesting to switch to optimistic synchronization.
When the number of rollbacks is excessively high, switching to conservative synchronization might be considered.
%And when neither of these two is an option, the simulation might just (temporarily) have to fall back to sequential simulation. % Falling back to sequential is not supported atm.

### 2016-06-30 publications/Journal/6-allocation.tex
Consistent italics and sans-serif fonts

Although the synchronization protocol is one of the defining factors in simulation performance, model allocation has a significant impact on which protocol is ideal.
Depending on the model structure, and how it is mapped to the different cores, it might not even make sense to parallelize at all.
Indeed, if the model is distributed such that frequent communication is necessary between cores, parallelism is naturally reduced.
This brings us to the topic of model allocation, as also implemented in PythonPDEVS~\cite{PythonPDEVS2}.

@@

\subsubsection{Sequential Simulation}
Since adevs does not use direct connection, we expect a noticable performance difference between dxex and adevs.
This is shown in Figure~\ref{fig:PHOLDtree_seq_dn_benchmark}, where the fanout ($n$) determines the performance penalty adevs suffers compared to dxex.
Profiling indeed indicates that an increase in width per subtree ($n$) leads to higher overheads in adevs due to the lack of direct connection.
Dxex uses direct connection, making it independent on fanout.
Performance of dxex is, in this model, only dependent on the number of models.
Slight deviations can still be seen, though, caused by the initialization overhead of direct connection.
Both adevs and dxex scale linearly in the number of atomic models.

>>>>

Although the synchronization protocol is one of the defining factors in simulation performance, model allocation has a significant impact on which protocol is ideal.
Depending on the model structure, and how it is mapped to the different cores, it might not even make sense to parallelize at all.
Indeed, if the model is distributed such that frequent communication is necessary between cores, parallelism is naturally reduced.
This brings us to the topic of model allocation, as also implemented in \textit{PythonPDEVS}~\cite{PythonPDEVS2}.

@@

\subsubsection{Sequential Simulation}
Since \textit{adevs} does not use direct connection, we expect a noticable performance difference between \textit{dxex} and \textit{adevs}.
This is shown in Figure~\ref{fig:PHOLDtree_seq_dn_benchmark}, where the fanout ($n$) determines the performance penalty \textit{adevs} suffers compared to \textit{dxex}.
Profiling indeed indicates that an increase in width per subtree ($n$) leads to higher overheads in \textit{adevs} due to the lack of direct connection.
\textit{Dxex} uses direct connection, making it independent on fanout.
Performance of \textit{dxex} is, in this model, only dependent on the number of models.
Slight deviations can still be seen, though, caused by the initialization overhead of direct connection.
Both \textit{adevs} and \textit{dxex} scale linearly in the number of atomic models.

### 2016-06-30 publications/Journal/7-related-work.tex
Added some more references to papers
Consistent italics and sans-serif fonts

In terms of code design and philosophy, dxex is most related to PythonPDEVS~\cite{PythonPDEVS}.
Performance of PythonPDEVS was still decent through the use of ``hints'' from the modeler.
In this spirit, we offer users the possibility to choose between different synchronization protocols.
This allows users to choose the most appropriate synchronization protocol, depending on the model.
Contrary to PythonPDEVS, however, dxex doesn't support distributed simulation~\cite{JDF}, model migrations~\cite{PythonPDEVS2}, or activity hints~\cite{PythonPDEVS_ACTIMS}.

Although PythonPDEVS offers very fast turnaround cycles, simulation performance was easily outdone by compiled simulation kernels.
In terms of performance, adevs~\cite{adevs} offered much faster simulation, at the cost of a significant compilation time.
The turnaround cycle in adevs is much slower though, specifically because the complete simulation kernel is implemented using templates in header files.
As a result, the complete simulation kernel has to be compiled again every time.
Similarly to vle~\cite{vle} and PowerDEVS~\cite{PowerDEVS}, dxex compromises by separating the simulation kernel into a shared library.
After the initial compilation of the simulation tool, only the model has to be compiled and linked to the shared library.
This significantly shortens the turnaround cycle, while still offering good performance.
In terms of performance, dxex is shown to be competitive with adevs.
Despite its high performance, adevs does not support optimistic synchronization, which we have shown to be highly relevant.

Previous \textsf{DEVS} simulation tools have already implemented multiple synchronization protocols, though none have done it in a strictly modular way that allows straightforward protocol switching for a single given model.
For example, adevs only supports conservative synchronization, and vle only supports experiment-level parallelism (\textit{i.e.}, run multiple experiments concurrently).
Closest to our support for multiple synchronization protocols is CD++~\cite{CD++}.
For CD++, both a conservative (CCD++~\cite{CCD++}) and optimistic (PCD++)~\cite{PCD++}) variant exist.
Despite the implementation of both protocols, they are different projects entirely.
Some features might therefore be implemented in CCD++ and not in PCD++, or vice versa.
And while this might not yet be that significant a problem to this day, this problem will only get worse when each project follows its own course.
Dxex, on the other hand, is a single project, where the choice of synchronization protocol is a simple configuration.
CD++, however, implements both conservative and optimistic synchronization for distributed simulation, whereas we limit ourselves to parallel simulation.
By limiting our approach to parallel simulation, we are able to achieve higher speedups through the use of shared memory communication.

% Paragraph runtime switching
In the PDES community, the problem of choosing between synchronization protocols is well known and documented~\cite{Jha:1994:UFC:195291.182480}.
The challenges of implementing such runtime switching have previously been explored already~\cite{Das:1996:APP:256562.256602}, and an implementation was given by, for example~\cite{perumalla2005musik}.
Our contribution entails bringing this same feature to the DEVS community, further expanding upon our unique feature of multiple synchronization protocols.

% Paragraph allocation effects
Model allocation and its impact on parallel performance has previously also been studied in the PDES community~\cite{PDESpartitioning}.
Referenced as partitioning of the simulation model, most studies distinguish between communication overhead and computational distribution (load balancing) as the two dimensions to partition over.
Partitioning a simulation model is identified as an issue to achieve scalability~\cite{Scalability}. 
Some research in the context of Parallel DEVS has been done, where they turn their attention to load balancing and communication overhead~\cite{PDEVSpartitioning}.
Our contribution studies the effect of partitioning with emphasis on the effect of communication between processes and in the presence of a flattened hierarchy. 
We focus on static partitioning since this is a limiting factor for our conservative synchronization implementation which does not support model relocation.
Model relocation, as implemented by PythonPDEVS~\cite{PythonPDEVS2}, might be an interesting addition to only model allocation at the start of simulation.

In summary, dxex tries to find the middle ground between the concepts of PythonPDEVS, the performance of adevs, and the multiple synchronization protocols of CD++.
To further profit from our multiple synchronization protocols in a single tool, we further added runtime switching between synchronization protocols and model allocation support.

>>>>

In terms of code design and philosophy, \textit{dxex} is most related to \textit{PythonPDEVS}~\cite{PythonPDEVS}.
Performance of \textit{PythonPDEVS} was still decent through the use of ``hints'' from the modeler.
In this spirit, we offer users the possibility to choose between different synchronization protocols.
This allows users to choose the most appropriate synchronization protocol, depending on the model.
Contrary to \textit{PythonPDEVS}, however, \textit{dxex} doesn't support distributed simulation~\cite{JDF}, model migrations~\cite{PythonPDEVS2}, or activity hints~\cite{PythonPDEVS_ACTIMS}.

Although \textit{PythonPDEVS} offers very fast turnaround cycles, simulation performance was easily outdone by compiled simulation kernels.
In terms of performance, \textit{adevs}~\cite{adevs} offered much faster simulation, at the cost of a significant compilation time.
The turnaround cycle in \textit{adevs} is much slower though, specifically because the complete simulation kernel is implemented using templates in header files.
As a result, the complete simulation kernel has to be compiled again every time.
Similarly to \textit{vle}~\cite{vle} and \textit{PowerDEVS}~\cite{PowerDEVS}, \textit{dxex} compromises by separating the simulation kernel into a shared library.
After the initial compilation of the simulation tool, only the model has to be compiled and linked to the shared library.
This significantly shortens the turnaround cycle, while still offering good performance.
In terms of performance, \textit{dxex} is shown to be competitive with \textit{adevs}.
Despite its high performance, \textit{adevs} does not support optimistic synchronization, which we have shown to be highly relevant.

Previous \textsf{DEVS} simulation tools have already implemented multiple synchronization protocols, though none have done it in a strictly modular way that allows straightforward protocol switching for a single given model.
For example, \textit{adevs} only supports conservative synchronization, and \textit{vle} only supports experiment-level parallelism (\textit{i.e.}, run multiple experiments concurrently).
Closest to our support for multiple synchronization protocols is \textit{CD++}~\cite{CD++}.
For \textit{CD++}, both a conservative (\textit{CCD++}~\cite{CCD++}) and optimistic (\textit{PCD++})~\cite{PCD++}) variant exist.
Despite the implementation of both protocols, they are different projects entirely.
Some features might therefore be implemented in \textit{CCD++} and not in \textit{PCD++}, or vice versa.
And while this might not yet be that significant a problem to this day, this problem will only get worse when each project follows its own course.
\textit{Dxex}, on the other hand, is a single project, where the choice of synchronization protocol is a simple configuration.
\textit{CD++}, however, implements both conservative and optimistic synchronization for distributed simulation, whereas we limit ourselves to parallel simulation.
By limiting our approach to parallel simulation, we are able to achieve higher speedups through the use of shared memory communication.

In the \textsf{PDES} community, the problem of choosing between synchronization protocols is well known and documented~\cite{Jha:1994:UFC:195291.182480}.
The challenges of implementing such runtime switching have previously been explored already~\cite{Das:1996:APP:256562.256602}, and an implementation was given by, for example~\cite{perumalla2005musik}.
Our contribution entails bringing this same feature to the \textsf{Parallel DEVS} community, further expanding upon our support for multiple synchronization protocols.

Model allocation and its impact on parallel performance has previously also been studied in the \textsf{PDES} community~\cite{PDESpartitioning}.
Referenced as partitioning of the simulation model, most studies distinguish between communication overhead and computational distribution (load balancing) as the two dimensions to partition over.
Partitioning a simulation model is identified as an issue to achieve scalability~\cite{Scalability}. 
Some research in the context of \textsf{Parallel DEVS} has been done, where they turn their attention to load balancing and communication overhead~\cite{PDEVSpartitioning}.
Our contribution studies the effect of partitioning with emphasis on the effect of communication between processes and in the presence of a flattened hierarchy. 
We focus on static partitioning since this is a limiting factor for our conservative synchronization implementation which does not support model relocation.
Model relocation, as implemented by \textit{PythonPDEVS}~\cite{PythonPDEVS2}, might be an interesting addition to only model allocation at the start of simulation.

In summary, \textit{dxex} tries to find the middle ground between the concepts of \textit{PythonPDEVS}, the performance of \textit{adevs}, and the multiple synchronization protocols of \textit{CD++}.
To further profit from our multiple synchronization protocols in a single tool, we further added runtime switching between synchronization protocols and model allocation support.

### 2016-06-30 publications/Journal/8-conclusions.tex
Consistent italics and sans-serif fonts

In this paper, we introduced DEVS-Ex-Machina (``dxex''), a new C++11-based \textsf{Parallel DEVS} simulation tool.
Our main contribution is the implementation of multiple synchronization protocols for parallel multicore simulation.
We have shown that there are indeed models which can be simulated significantly faster using either synchronization protocol.
Dxex allows the user to choose between either conservative or optimistic synchronization at the start of simulation.
Depending on observed model behaviour and simulation performance, runtime switching between synchronization protocols can be used.

Notwithstanding our modularity, dxex achieves performance competitive to adevs, another very efficient \textsf{DEVS} simulation tool.
Performance is measured both in elapsed time, and memory usage.
Our empirical analysis shows that allocation of models over kernels is critical to enable a parallel speedup. Furthermore we have shown when and why optimistic synchronization can outperform conservative and vice versa.

Future work is possible in several directions.
First, our implementation of optimistic synchronization should be more tolerant to low-memory situations.
In its current state, simulation will simply halt with an out-of-memory error.
Having simulation control, which can throttle the speed of nodes that use up too much memory, has been shown to work in these situations~\cite{FujimotoBook}.
Faster GVT implementations~\cite{Fujimoto:1997:CGV:268403.268404,Bauer:2005:SND:1069810.1070159} might further help to minimize this problem.
Second, the runtime switching between synchronization protocols can be driven using machine learning techniques.
The simulation engine is already capable of collecting data to inform such a process, and is designed to listen for commands from an external component.
Third, automatic allocation might be possible by analysis of the connections between models.
This information is already used in dxex to construct the dependency graph in conservative synchronization.
A graph algorithm that distributes models, while avoiding cycles, could be used to offer a parallel speedup in either optimistic or conservative synchronization.
Similarly, it could serve as a default parallel allocation scheme that can be improved by the user.

>>>>

In this paper, we introduced \textit{DEVS-Ex-Machina} (``\textit{dxex}''), a new C++11-based \textsf{Parallel DEVS} simulation tool.
Our main contribution is the implementation of multiple synchronization protocols for parallel multicore simulation.
We have shown that there are indeed models which can be simulated significantly faster using either synchronization protocol.
\textit{Dxex} allows the user to choose between either conservative or optimistic synchronization at the start of simulation.
Depending on observed model behaviour and simulation performance, runtime switching between synchronization protocols can be used.

Notwithstanding our modularity, \textit{dxex} achieves performance competitive to \textit{adevs}, another very efficient \textsf{DEVS} simulation tool.
Performance is measured both in elapsed time, and memory usage.
Our empirical analysis shows that allocation of models over kernels is critical to enable a parallel speedup. Furthermore we have shown when and why optimistic synchronization can outperform conservative and vice versa.

Future work is possible in several directions.
First, our implementation of optimistic synchronization should be more tolerant to low-memory situations.
In its current state, simulation will simply halt with an out-of-memory error.
Having simulation control, which can throttle the speed of nodes that use up too much memory, has been shown to work in these situations~\cite{FujimotoBook}.
Faster GVT implementations~\cite{Fujimoto:1997:CGV:268403.268404,Bauer:2005:SND:1069810.1070159} might further help to minimize this problem.
Second, the runtime switching between synchronization protocols can be driven using machine learning techniques.
The simulation engine is already capable of collecting data to inform such a process, and is designed to listen for commands from an external component.
Third, automatic allocation might be possible by analysis of the connections between models.
This information is already used in \textit{dxex} to construct the dependency graph in conservative synchronization.
A graph algorithm that distributes models, while avoiding cycles, could be used to offer a parallel speedup in either optimistic or conservative synchronization.
Similarly, it could serve as a default parallel allocation scheme that can be improved by the user.
